---
title: "Project 2"
author: "Natalia Sifnugel"
date: "2023-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytuesdayR)
library(here)
library(lubridate)
library(sessioninfo)

```

# Part 1 Fun with functions

### Part 1A: Exponential transformation

```{r}

Exp <- function(x, k) {
  exp = 1 
 for(i in 1:k){
   exp2 = x^i/factorial(i)
   exp = exp + exp2
 } 
  return(exp)
}

```

#### Sample outputs

```{r}

Exp(2,4)

# check 

1 + 2 + (2^2/factorial(2)) + (2^3/factorial(3)) + (2^4/factorial(4))

```

### Part 1B: Sample mean and sample standard deviation

```{r}

sample_mean <- function(x) {
  numerator = 0
  denominator = 0
  for(i in seq_along(x)){
    numerator = numerator + x[i] 
    denominator = denominator + 1
  } 
  mean = numerator/denominator
  return(mean)
}

sample_sd <- function(x) {
  numerator = 0
  denominator = 0
  for(i in seq_along(x)){
    numerator = numerator + (x[i]-sample_mean(x))^2
    denominator = denominator + 1
  } 
  sd = sqrt(numerator/(denominator-1))
  return(sd)
}

```

#### Sample outputs

```{r}
x = c(1,22,14,67,10)

sample_mean(x)

# check 
mean(x)

sample_sd(x)

# check 
sd(x)
```

### Part 1C: Confidence intervals

```{r}

calculate_CI <- function(x, conf = 0.95) {
  alpha <- 1 - conf
  degrees_freedom <- length(x) - 1
  t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE) 
  lower_tail = sample_mean(x) - t_score*(sample_sd(x)/sqrt(length(x)))
  upper_tail = sample_mean(x) + t_score*(sample_sd(x)/sqrt(length(x)))
  CI = c(lower_tail,upper_tail)
  return(CI)
}

```

#### Sample outputs

```{r}
x = c(1,22,14,67,10)

calculate_CI(x,0.90)
calculate_CI(x,0.80)

# check 

dat <- data.frame(x = x)
fit <- lm(x ~ 1, dat)
confint(fit, level = 0.90)

confint(fit, level = 0.80)

```

# Part 2: Wrangling data

### Downloading datasets
```{r}

if (!dir.exists(here("data"))) {
  dir.create(here("data"))
}

if (!file.exists(here("data", "tuesdata_rainfall.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-01-07")
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature

    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = here("data", "tuesdata_rainfall.RDS"))
    saveRDS(tuesdata$temperature, file = here("data", "tuesdata_temperature.RDS"))
}

rainfall <- readRDS(here("data", "tuesdata_rainfall.RDS"))
temperature <- readRDS(here("data", "tuesdata_temperature.RDS"))

```

### Wrangling rainfall dataset and joining it to temperature dataset
```{r}
rainfall_temp <-rainfall %>%
  drop_na()  %>%
  mutate(date = ymd(paste(year,month,day, sep = "-")), city_name = toupper(city_name)) %>%
  select(-month,-day) %>%
  inner_join(temperature, by = c("date","city_name"), relationship = "many-to-many")

show(rainfall_temp)
```

# Part 3: Data visualization

### Part 3A: Plotting temperature data over time

```{r fig.height=14, fig.width=18}
rainfall_temp %>%
  filter(year >= 2014) %>%
  ggplot() +
  geom_line(aes(x = date, y = temperature, color = temp_type))+
  facet_grid(city_name ~. ) +
  labs(title = "Temperature vs. Year",
       subtitle = "Canberra experiences greatest range in daily temperature. In Sydney, peak daily temperatures are rising.",
       caption = "Source: Australian Government, Bureau of Meteorology ",
      x = "Date",
      y = "Temperature (C)") +
  theme_light() +
  theme(axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        plot.title = element_text(size = 22),
        plot.subtitle = element_text(size = 16))
```

### Part 3B: Plotting rainfall over time

```{r fig.height=8, fig.width=14}

rainfall_histogram <- function(x, y) {
  if(x %in% rainfall_temp$city_name){
    rainfall_city <- rainfall_temp %>%
    filter(city_name == x)
  } else {
    stop(x," is not in the dataset.")
  }
  if(y %in% rainfall_city$year){
      rainfall_city %>%
      filter(year == y) %>%
      ggplot(aes(log(rainfall + 0.000000000000001))) +
        geom_histogram(fill = 'lightblue', color = 'black') +
        labs(title = paste("Distribution of Rainfall in",x,"in",y, sep = " "),
                subtitle =  paste("Use this histogram to determine if",y,"was a year of deluge, drought, or expected rain in",x,sep = " "),
                caption = "Source: Australian Government, Bureau of Meteorology ",
                x = "log(Rainfall in mm)",
                y = "Number of Days") +
        theme_light() +
        theme(axis.title.x = element_text(size = 12),
              axis.title.y = element_text(size = 12),
              plot.title = element_text(size = 18),
              plot.subtitle = element_text(size = 12))
  } else{
    stop(y," not in dataset for ",x,".")
  }
}
```
#### Function design 

In this function, I used an if statement to check if the city is in the data set. If the city is in the data set, then a new filtered data frame is created in which only the values for the specified city are included. This data set then enters a new if else statement to check if the year (second input) exists for this city. If the year does exist in the filtered city data set, then a histogram is generated for that city for that year. If these two conditions were not met, the function returns a statement indicating the data are not present. 

Because the log(0) is an infinite number, days with 0 mm of rainfall were initially dropped from the histogram. Days of 0 rainfall for this data set as a whole are important because it seeks to understand the relationship between temperate, rain, and fires. So, to ensure that days of 0 rainfall are included, I added a small constant (rainfall + 0.000000000000001). 


#### Example output

```{r fig.height=8, fig.width=14}

rainfall_histogram('PERTH',2000)

rainfall_histogram('BRISBANE',1981)

rainfall_histogram('SYDNEY',2014)

```

# Part 4: Apply functions and plot

### Part 4A: Summary of rainfall data for each city since 2014

```{r}

rain_df <- rainfall_temp %>%
  filter(year >= 2014) %>%
  group_by(city_name, year) %>%
  summarize(tibble(
    `mean` = sample_mean(rainfall),
    `sd` = sample_sd(rainfall),
    `lower_bound` = calculate_CI(rainfall,0.95)[1],
    `upper_bound` = calculate_CI(rainfall,0.95)[2]
  )
)

show(rain_df)
```

### Part 4B: Plotting mean rainfall 

```{r fig.height=8, fig.width=20}

rain_df %>%
  ggplot() +
  geom_point(aes(year,mean, color = city_name)) +
  geom_line(aes(year,mean, color = city_name)) +
  facet_grid(.~ city_name) +
  geom_errorbar(aes(ymin= lower_bound, ymax= upper_bound, x = year), width = 0.2) +
  labs(title = "Mean Rainfall by Year",
       subtitle =  "Average rainfall is on the decline across major cities in Australia.",
       caption = "Source: Australian Government, Bureau of Meteorology ",
       x = "Year",
       y = "Rainfall in mm") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        plot.title = element_text(size = 22),
        plot.subtitle = element_text(size = 16))

```


# R session information

```{r}

options(width = 120)
sessioninfo::session_info()

```
